---
title: "Practical Machine Learning Course Project"
author: "Felix Schneider"
date: "7/31/2020"
output: 
  html_document: 
    toc: yes
    number_sections: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	include = FALSE
)
knitr::opts_chunk$set(collapse = TRUE)
library(dplyr)
library(ggplot2)
library(caret)
library(doParallel)
library(randomForest)
library(e1071)
library(foreach)
library(import)
```

# Introduction
The purpose of the analysis is to use accelerometer data on belt, arm, forearm and dumbbell of 6 individuals performing barbell lifts. They are asked to perform the lifts correctly as well as incorrectly in 4 different ways. The goal of the analysis is to predict the manner in which they performed the exercise.

# The Data
The data come from this source: http://groupware.les.inf.puc-rio.br/har
```{r download_and_load_data, cache=TRUE, include=TRUE, message=FALSE}
trainURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(trainURL,"pml-training.csv")
download.file(testURL, "pml-testing.csv")
pml_training <- read.csv("pml-training.csv",na.strings=c("NA","#DIV/0!"))
pml_testing  <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!"))
```
Check the dimensions.
```{r dim_training_testing, include=TRUE}
dim(pml_training)
dim(pml_testing)
```
Check for `NA`s.
```{r check_for_NAs, include=TRUE}
table(colSums(is.na(pml_training)))
table(colSums(is.na(pml_testing)))
```
There are only 60 columns that have no `NA`s. 67 columns have 19216 `NA`s. This can be seen by checking the `new_window` variable.
```{r check_new_window, include=TRUE}
table(pml_training$new_window)
table(pml_testing$new_window)
```
Remove the covariates that are `NA` in the pml_**testing** set and call it `quiz` set, because we want to use those covariates for training that are provided for prediction in the Course Project Prediction Quiz. 
```{r remove_NA_columns, include=TRUE}
notNA <- colSums(is.na(pml_testing))==0
train_notNA <- pml_training[,notNA][,-1]
quiz  <- pml_testing[,notNA][,-1]
```

# The Features
## Measured Variables
The label `classe` assumes one of 5 distinct values.
```{r unique_train_classe, include=TRUE}
unique(train_notNA$classe)
```
The labels correspond to the following executions of the weight lifting exercise.
 
 label | execution
 ------|----------
 A | correct execution
 B | throwing elbows to the front
 C | lifting dumbbell only halfway
 D | lowering doumbbell only halfway
 E | throwing the hips to the front
 
The variables come from 4 sensors (IMUs, Inertial Measurement Unit) mounted on

1. arm
1. forearm
1. belt
1. dumbbell

of the participants.

Each IMU measured the 3 cartesian components (`x`,`y`,`z`) of three different quantities:

1. linear acceleration (`accel`)
1. angular acceleration (`gyros`)
1. earth magnetic field (`magnet`)

Further, the orientation of the IMU was measured with the 3 angles

1. roll
1. pitch
1. yaw

Finally, 8 derived quantities are calculated

1. mean (`avg`)
1. variance (`var`)
1. standard deviation (`stddev`)
1. minimum (`min`)
1. maximum (`max`)
1. amplitude (`amplitude`)
1. kurtosis (`kurtosis`)
1. skewness (`skewness`)

## Features
```{r prepare_train_dataset, include=TRUE}
train <- train_notNA %>%
  select(roll_belt:magnet_forearm_z,classe) %>%
  mutate(classe=factor(classe))
X_quiz  <- select(quiz,roll_belt:magnet_forearm_z)
```

# Algorithm
## Model
- *How do I build the model?*
- *How do I use cross validation?*
- *What is the expected out-of-sample error?*
- *Why did I make the choices?*

The problem is a *multiclass* classification problem.
Start with a coarse model to get a feel of the calculation time.

Set aside an evaluation data set to evaluate out-of-sample error.
```{r create_data_partitions}
inBuild <- createDataPartition(y=train$classe, p=0.8, list=FALSE)
build <- train[ inBuild,]
evalu <- train[-inBuild,]
```

Set resampling method to `"cv"` for cross validation.
```{r set_traincontrol_parameters}
tc <- trainControl(method="cv")
```

I chose the train method `parRF` because I want to plot Variable Importance and I hope this method provides this because it requires the package `RandomForest` which creates a model object that can be plotted with `varImpPlot()`.
```{r train_model, cache=TRUE}
cl <- makePSOCKcluster(8)
registerDoParallel(cl)
tic <- Sys.time()
mdl02parRF <- train(classe~., data=build, method="parRF", trControl=tc, importance=TRUE)
toc <- Sys.time()
stopCluster(cl)
print(toc-tic)
mdl02parRF$times
```
```{r print_and_save_model}
mdl02parRF
save(mdl02parRF,file="./RData/mdl02RF.RData")
```
## Cross Validation

## Out-Of-Sample Error
```{r}
X_evalu <- select(evalu,-classe)
confusionMatrix(predict(mdl02parRF,X_evalu),evalu$classe)
```

# Prediction
```{r}
prd02parRF <- predict(mdl02parRF,X_quiz)
prd02parRF
save(prd02parRF,file="./RData/prd02parRF.RData")
```

# Conclusion